\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{Acknowledgements}

This work is indebted to traditions of analytical exposition exemplified by Edward Tufte, whose principles of visual clarity and data-driven argument have shaped every page. It is equally indebted to historians of science who treat instruments, calculations, and error not as peripheral details but as primary actors in historical change. These traditions—that precision is inseparable from clarity, and that material evidence speaks louder than institutional narratives—have guided my approach throughout.

The archivists and curators of the Royal Observatory, Greenwich, and the National Maritime Museum generously provided access to manuscripts, observation logs, correspondence, and instruments that form the material foundation of this work. Their stewardship of these collections, and their willingness to contextualize rather than merely preserve them, transformed what might have been dusty records into living evidence of how precision was built and maintained. I am similarly grateful to the staff of Cambridge University Library, Trinity College, Cambridge, and the British Library, whose assistance illuminated the broader intellectual landscape within which Greenwich Observatory operated.

Several colleagues and readers generously worked through the mathematical derivations, challenged interpretations, and caught errors in both calculation and argument. Their skepticism was exactly what historical rigor demanded. In particular, I thank those who insisted on understanding not just \textit{what} was computed, but \textit{how}—the difference between narrative and history. Remaining errors, whether of fact or judgment, are mine alone.

This book was composed using LaTeX and the memoir document class, tools that enabled the integration of text, mathematics, tables, and figures as a unified analytical argument. The choice of these tools reflects the book's fundamental commitment to precision in presentation. I am grateful to the open-source communities that maintain these ecosystems, often unseen and always underappreciated.

I thank those close to me for tolerating long digressions into angles, clocks, and 18th-century tables, and for providing the sustained quiet necessary to think clearly about difficult problems.

\vspace{0.5\baselineskip}

Any errors of fact, calculation, interpretation, or judgment are solely my own.
